---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# tidytableone

<!-- badges: start -->
<!-- badges: end -->

> Tidy construction of "Table 1" study-characteristic summaries for biomedical research.

`tidytableone` produces the descriptive "Table 1" that opens nearly every clinical paper: a stratified summary of study characteristics with appropriate continuous and categorical summaries, statistical tests, and missing-data handling. The output comes in two forms — a long tidy tibble for programmatic use, or a polished publication-style table for direct inclusion in reports.

## Installation

```r
# install.packages("pak")
pak::pak("emilelatour/tidytableone")
```

## Usage

The two-function workflow is `create_tidytableone()` → `adorn_tidytableone()`:

- `create_tidytableone()` — computes summaries, tests, and standardized mean differences; returns a long tidy tibble of results.
- `adorn_tidytableone()` — turns that tidy tibble into a publication-ready wide table with formatted cells, indented level rows, and a single p-value column.

```{r usage-basic, message = FALSE}
library(tidytableone)

tidy_t1 <- create_tidytableone(
  data   = pbc_mayo,
  strata = "trt",
  vars   = c("age", "sex", "ascites", "hepato", "spiders", "edema",
             "bili", "albumin", "stage")
)

adorn_tidytableone(tidy_t1)
```

For an unstratified table, omit `strata`:

```{r usage-no-strata}
create_tidytableone(
  data = pbc_mayo,
  vars = c("age", "sex", "bili")
) |>
  adorn_tidytableone()
```

## Checkbox-style multi-response variables

REDCap-style "check all that apply" fields are common in clinical data and awkward to summarize with most table-one packages. `tidytableone` handles them as first-class citizens. A small synthetic example:

```{r checkbox-data, message = FALSE}
library(tibble)
library(dplyr)

set.seed(42)
n <- 200

cb_data <- tibble(
  treatment = sample(c("Drug", "Placebo"), n, replace = TRUE),
  age       = round(rnorm(n, 55, 12)),
  race___1  = sample(c("Checked", "Unchecked"), n, replace = TRUE, prob = c(0.55, 0.45)),
  race___2  = sample(c("Checked", "Unchecked"), n, replace = TRUE, prob = c(0.20, 0.80)),
  race___3  = sample(c("Checked", "Unchecked"), n, replace = TRUE, prob = c(0.10, 0.90)),
  race___98 = sample(c("Checked", "Unchecked"), n, replace = TRUE, prob = c(0.05, 0.95))
)

checkbox_spec <- tribble(
  ~var,        ~overall_lbl, ~checkbox_lbl,
  "race___1",  "Race",       "White",
  "race___2",  "Race",       "Black or African-American",
  "race___3",  "Race",       "American Indian or Alaska Native",
  "race___98", "Race",       "Prefer not to answer"
)

create_tidytableone(
  data     = cb_data,
  strata   = "treatment",
  vars     = c("age", "race___1", "race___2", "race___3", "race___98"),
  checkbox = checkbox_spec
) |>
  adorn_tidytableone()
```

The block renders with a single labeled header ("Race, More than one response allowed") followed by indented per-option rows, with group-based percentages (denominator = stratum N) and per-level p-values. Multiple blocks per table are supported, including blocks whose member variables don't share a common stem.

## Statistical tests

`create_tidytableone()` computes every candidate p-value; `adorn_tidytableone()` picks one per variable for display:

| Variable type | Default test | Override options |
|---|---|---|
| Continuous, 2 groups | Welch's t-test | `equal_variance = TRUE` → Student's t; `nonnormal` → Wilcoxon rank-sum |
| Continuous, 3+ groups | Welch's ANOVA | `equal_variance = TRUE` → one-way ANOVA; `nonnormal` → Kruskal-Wallis |
| Categorical | Chi-squared | `exact` → Fisher's exact; `monte_carlo_p` → simulated p-values |

The tidy tibble contains every test variant (Welch's, Student's, Kruskal, Bartlett, Levene, chi-squared with and without continuity correction, Fisher's, and simulated versions), so you can switch displays without recomputing.

## Output

`create_tidytableone()` returns a long tibble with one row per (stratum × variable × level) combination. Key columns:

| Column | What it is |
|---|---|
| `strata_var`, `strata` | Stratification variable name and level (`"Overall"` row included) |
| `var`, `level`, `label` | Variable name, factor level (for categorical), display label |
| `var_type`, `class` | `"continuous"` / `"categorical"`; `"checkbox"` flag for multi-response |
| `n`, `complete`, `missing` | Counts |
| `mean`, `sd`, `p0`–`p100`, `cv` | Continuous summaries |
| `n_level`, `n_strata`, `pct` | Categorical counts and percentages |
| `chisq_test`, `fisher_test`, `oneway_test_*`, `kruskal_test`, … | Test p-values |
| `smd` | Standardized mean difference |

`adorn_tidytableone()` returns a wide tibble suitable for direct printing or piping into `gt`, `flextable`, `kableExtra`, etc.

## What this isn't

`tidytableone` is one of several R packages for descriptive tables; alternatives include [`tableone`](https://CRAN.R-project.org/package=tableone), [`gtsummary`](https://www.danieldsjoberg.com/gtsummary/), and [`compareGroups`](https://CRAN.R-project.org/package=compareGroups), each with different design priorities. Pick what fits your workflow.

## License

MIT