---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidytableone

<!-- badges: start -->
<!-- badges: end -->

The goal of `tidytableone` is to enable a tidy approach to creating the "Table 1" of study characteristics common in bimoedical research.

## Installation

You can install the released version of tidytableone from [CRAN](https://CRAN.R-project.org) with:

``` r
# install.packages("devtools")
devtools::install_github("emilelatour/tidytableone)
```

## Example

Load the packages for the example
```{r example}
library(tidytableone)
library(tidyverse)
library(knitr)

```

Take a glimpse at the data that comes in the `tidytableone` package:

```{r}
dplyr::glimpse(pbc_mayo)
```

Use the main function of this package to make a tidy table one

```{r}
tab1 <- tidytableone::create_tidy_table_one(data = pbc_mayo, 
                                            strata = "trt", 
                                            .vars = c("time", "status", "trt", "age", "sex", "ascites", "hepato",
                                                      "spiders", "edema", "bili", "chol", "albumin", "copper", "alk_phos",
                                                      "ast", "trig", "platelet", "protime", "stage"))


```

Use the `dplyr::glimpse` command to see the tidy table one

```{r}
dplyr::glimpse(tab1)
```

Some information is only available for categorical variables and some information for continous variables.

Then with some simple data manipulation we can reshape to make a table one...

```{r}
tab1 %>% 
  mutate(strata = forcats::fct_inorder(strata), 
         n_pct = scales::percent(n_strata / n_level), 
         mean = scales::number(mean, accuracy = 0.1), 
         value = dplyr::if_else(mean == "NA", n_pct, mean), 
         p_value = dplyr::coalesce(oneway_test, chisq_test), 
         p_value = scales::pvalue(p_value)) %>% 
  dplyr::distinct(strata, 
                var, 
                level, 
                value, 
                p_value) %>% 
  tidyr::spread(., 
                key = strata, 
                value = value) %>% 
  dplyr::select(-p_value, dplyr::everything(), p_value) %>% 
  knitr::kable()
  
```

This is rough, but you get the idea. The plan is to next create a function to make the table based on specifications.
